---
title: "Headspace sample processing Part 2"
output: 
  html_document:
    toc: yes
    toc_float:
      collapsed: false
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE, message=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(knitr)
library(readr)
library(dplyr)
library(RPostgreSQL)
library(uuid)
library(tidyr)
library(lubridate)
library(magrittr)
library(ggplot2)
library(RColorBrewer)
```

```{r connect, include=FALSE, message=FALSE}
password <- scan(".pgpass", what="")
db <- dbConnect(PostgreSQL(), 
                host = "sesync-postgis01.research.sesync.org",
                dbname = "choptank", 
                user = "palmergroup",
                password = password)
```

# Description

Calculate dissolved gas concentrations for headspace samples from PPM data. Then insert dissolved gas concentration data into database.

# Data

* Get PPM data from database for source air and samples
* ppmv data for source gas - determined from GC workflow
* ppmv data for sample gas - determined from GC workflow
* water temperature - measured in the field or offset from GSODR
* barometric pressure - measured in the field or from GSODR data
* source gas sample volume - default 20 mL
* sample gas sample volume - default 40 mL

**PPM data**

Reshape the data so there is one row for each headspace sample with the required inputs for the `def.calc.sdg()` function. 

```{r, include=FALSE}
source('R/db_get_hs_sources.R')
```

```{r}
hs_sources <- db_get_hs_sources() 
hs_sources %>% kable()
```

Get the methane and co2 ppm data for each sample 

```{r}
db_get_gc_data <- function(){
sql <- "SELECT sf.samplingfeaturecode, v.variablecode, u.unitsname, mrv.datavalue, mrv.valuedatetime  
  FROM odm2.measurementresultvalues mrv, odm2.results r, odm2.variables v, odm2.units u, odm2.samplingfeatures sf, odm2.featureactions fa, odm2.methods m, odm2.actions act
 WHERE r.variableid = v.variableid
 AND fa.actionid = act.actionid
 AND act.methodid = m.methodid
 AND r.featureactionid = fa.featureactionid
 AND fa.samplingfeatureid = sf.samplingfeatureid
 AND r.unitsid = u.unitsid
 AND mrv.resultid = r.resultid 
 AND r.sampledmediumcv = 'gas'
 AND m.methodcode = 'GC'"

sql <- gsub("\n", "", sql)
dbGetQuery(db, sql)
}
```
```{r}
ppms <- db_get_gc_data()
ppms %>% kable()
```

Now we have hs sources and the ppm data. Join in wide format for the sdg function

```{r}
sdg_data <- hs_sources %>% 
  left_join(ppms, by = c("source" = "samplingfeaturecode")) %>%
  dplyr::select(-unitsname, -valuedatetime) %>%
  filter(variablecode %in% c("carbonDioxide", "methane")) %>%
  spread(key = variablecode, value = datavalue) %>%
  rename(concentrationCO2Air = carbonDioxide, concentrationCH4Air = methane) %>%
  left_join(ppms, by = c("hs" = "samplingfeaturecode")) %>%
  dplyr::select(-unitsname) %>%
  filter(variablecode %in% c("carbonDioxide", "methane")) %>%
  spread(key = variablecode, value = datavalue) %>%
  rename(concentrationCO2Gas = carbonDioxide, concentrationCH4Gas = methane)
```

get table of where each sample was collected 

```{r, warning=FALSE}
sql <- "SELECT * FROM odm2.relatedfeatures WHERE relationshiptypecv = 'wasCollectedAt'"
sql <- gsub("\n", "", sql)
rf <- dbGetQuery(db, sql)
sampfeatures <- dbReadTable(db, c("odm2", "samplingfeatures"))
samps_join_sites <- rf %>% 
  left_join(sampfeatures, by = c("samplingfeatureid" = "samplingfeatureid")) %>%
  dplyr::select(relationid, samplingfeaturecode, relationshiptypecv, relatedfeatureid) %>%
  rename(specimencode = samplingfeaturecode) %>%
  left_join(sampfeatures, by = c("relatedfeatureid" = "samplingfeatureid")) %>%
  dplyr::select(specimencode, relationshiptypecv, samplingfeaturecode)
kable(samps_join_sites)
```

add sites to sdg data

```{r}
sdg_data %<>% left_join(samps_join_sites, by = c("hs" = "specimencode")) %>%
  dplyr::select(- relationshiptypecv) %>% rename(site = samplingfeaturecode)
```

**Water Temp**

add in water temp data from field data collection. 

**Barometric pressure**

Add column for barometric pressure and temp on day of sampling from Dover AFB GSOD data. 

```{r, message=FALSE}
gsod <- read_csv("data/gsod_data.csv")
```

convert gsod data from mb to kPa

```{r}
gsod$STP_kPa <- gsod$STP*0.1
gsod$TEMP_C <- gsod$TEMP
```


relevant columns in gsod data are YEARMODA, STP_kPa, TEMP. TEMP is mean daily temperature in Celcius (GSOD package does conversion to SI units).

```{r}
sdg_data$date <- date(sdg_data$valuedatetime)
gsod %<>% dplyr::select(YEARMODA, STP_kPa, TEMP_C)
sdg_data %<>% left_join(gsod, by = c("date" = "YEARMODA"))
```

```{r}
sdg_data$concentrationN2OAir = NA
sdg_data$concentrationN2OGas = NA
sdg_data %<>% mutate(headspaceTemp = TEMP_C) %>% 
  rename(barometricPressure = STP_kPa, waterTemp = TEMP_C) %>%
  mutate(gasVolume = 20, waterVolume = 40)
```

# Calculations

Calculations are based on the NEON dissolved gas function. Add proper citation to [here](https://github.com/NEONScience/NEON-dissolved-gas). 

```{r}
source('def.calc.sdg.R')
```

Use function by supplying column names and defaults. Output is concentrations and 100% saturation concentrations as additional columns on the data frame. 

```{r}
sdg_out <- def.calc.sdg(inputFile = sdg_data)
```

output is in [M]

```{r}
sdg_results <- sdg_out %>%
  mutate(dissolvedCH4_umol = dissolvedCH4 * 1e6,
         dissolvedCO2_umol = dissolvedCO2 * 1e6,
         satCH4_umol = satCH4 * 1e6) %>%
  dplyr::select(hs, site, date, valuedatetime, dissolvedCH4_umol, dissolvedCO2_umol, satCH4_umol)
```

# Plots 

woo data!

```{r}

# make the hs id a factor with the order corresponding to what site the samples are from
sdg_results$hs_factor <- factor(sdg_results$hs, ordered = TRUE, levels = sdg_results[order(sdg_results$site),"hs"])

sdg_results %>%
  gather(key = gas, value = umol_per_L, dissolvedCH4_umol:satCH4_umol) %>%
  filter(gas %in% c("dissolvedCH4_umol", "dissolvedCO2_umol")) %>%
  ggplot(aes(x = hs_factor, y = umol_per_L, fill = site)) +
  geom_bar(stat = "identity", col = "black", position = position_dodge()) +
  facet_wrap(~gas, scales = "free") +
  coord_flip() +
  theme_bw() +
  scale_fill_brewer(palette = "Set1")

```

```{r}
sdg_results %>%
  ggplot(aes(x = dissolvedCH4_umol, y = dissolvedCO2_umol, fill = site)) +
  geom_point(pch = 21, col = "black", size = 4) +
  theme_bw() + 
  scale_fill_brewer(palette = "Set1")
  # scale_y_log10() + scale_x_log10()
```

```{r}
sdg_results %>%
  ggplot(aes(x = dissolvedCH4_umol, y = dissolvedCO2_umol + dissolvedCH4_umol, fill = site)) +
  geom_point(pch = 21, col = "black", size = 4) +
  theme_bw() + 
  scale_fill_brewer(palette = "Set1")
```

Reshape concentration data to put values in database for each sample

# Insert concentration results

Look at the data and assess for quality - add marginal quality code to samples that seem to not be right. 

```{r}
sdg_results$qualitycode <- "unknown"
flag_marginal <- c("KLH Feb18xx107", "KLH Feb18xx102", "KLH 596", "KLH 584", "KLH 595", "KLH 504")
sdg_results[sdg_results$hs %in% flag_marginal, "qualitycode"] <- "marginal"
```

check for variables and units

```{r, include=FALSE}
source('R/db_add_new_variable.R')
```

```{r addvars, eval=FALSE}
db_add_new_variable(variabletypecv = "WaterQuality", variablecode = "carbonDioxideDissolved", 
                    variablenamecv = "Carbon Dioxide, dissolved", variabledefinition = "Dissolved Carbon dioxide (CO2)")
```

```{r}
dbGetQuery(db, "SELECT * FROM odm2.units WHERE unitsname = 'Micromole per Liter'")
```


```{r, eval=FALSE}
source('R/db_add_method.R')
db_add_method(methodtypecv = "derivation",
              methodcode = "calc_sdg",
              methodname = "Calculate dissolved gas concentration",
              methoddescription = "Calculate dissolved gas concentration in surface water based on headspace equilibration data")
```

```{r}
utcoffset <- format(Sys.time(), "%z")
utcoffset <- as.integer(substr(utcoffset, 1,3))
```

**result metadata**

* Measurement doesn't quite seem like the correct type of result for derived results.
* Processling level code 3 for an "interpreted product"
* sampled medium is the dissolved gas so liquid aqueous? 

```{r}

db_insert_sdg_results <- function(x){
  sql_blanks <- 'WITH
  newact AS (
  INSERT INTO odm2.actions (
    actiontypecv, methodid, begindatetime, begindatetimeutcoffset)
  VALUES (
    \'%s\', 
    (SELECT methodid FROM odm2.methods WHERE methodcode = \'%s\'), 
    \'%s\', 
    \'%s\')
  RETURNING actionid),

  newfa AS (
  INSERT into odm2.featureactions (
    samplingfeatureid, actionid)
  VALUES (
    (SELECT samplingfeatureid FROM odm2.samplingfeatures WHERE samplingfeaturecode = \'%s\'),
    (SELECT newact.actionid FROM newact))
  RETURNING featureactionid),

  newresult AS (
  INSERT INTO odm2.results (
    featureactionid, resultuuid, resulttypecv, variableid, unitsid, processinglevelid, sampledmediumcv, valuecount)
  VALUES (
    
    (SELECT newfa.featureactionid FROM newfa),
    \'%s\',
    \'%s\',
    (SELECT variableid FROM odm2.variables WHERE variablecode = \'methaneDissolved\'),
    (SELECT unitsid FROM odm2.units WHERE unitsname = \'Micromole per Liter\'),
    \'%s\',
    \'%s\',
    \'%s\'),
  (
    (SELECT newfa.featureactionid FROM newfa),
    \'%s\',
    \'%s\',
    (SELECT variableid FROM odm2.variables WHERE variablecode = \'carbonDioxideDissolved\'),
    (SELECT unitsid FROM odm2.units WHERE unitsname = \'Micromole per Liter\'),
    \'%s\',
    \'%s\',
    \'%s\')
  RETURNING resultid, variableid),

  newmr1 AS (
  INSERT INTO odm2.measurementresults (
    resultid,
    censorcodecv,
    qualitycodecv,
    aggregationstatisticcv,
    timeaggregationinterval,
    timeaggregationintervalunitsid)
  VALUES (
    (SELECT newresult.resultid FROM newresult WHERE variableid = (SELECT variableid FROM odm2.variables WHERE variablecode = \'methaneDissolved\')),
    \'%s\',
    \'%s\', 
    \'%s\', 
    \'%s\', 
    \'%s\')),

  newmrv1 AS (
  INSERT INTO odm2.measurementresultvalues (
    resultid,
    datavalue,
    valuedatetime,
    valuedatetimeutcoffset)
  VALUES (
    (SELECT newresult.resultid FROM newresult WHERE variableid = (SELECT variableid FROM odm2.variables WHERE variablecode = \'methaneDissolved\')),
    \'%s\',
    \'%s\', 
    \'%s\')),

  newmr3 AS (
  INSERT INTO odm2.measurementresults (
    resultid,
    censorcodecv,
    qualitycodecv,
    aggregationstatisticcv,
    timeaggregationinterval,
    timeaggregationintervalunitsid)
  VALUES (
    (SELECT newresult.resultid FROM newresult WHERE variableid = (SELECT variableid FROM odm2.variables WHERE variablecode = \'carbonDioxideDissolved\')),
    \'%s\',
    \'%s\', 
    \'%s\', 
    \'%s\', 
    \'%s\'))

  INSERT INTO odm2.measurementresultvalues (
    resultid,
    datavalue,
    valuedatetime,
    valuedatetimeutcoffset)
  VALUES (
    (SELECT newresult.resultid FROM newresult WHERE variableid = (SELECT variableid FROM odm2.variables WHERE variablecode = \'carbonDioxideDissolved\')),
    \'%s\',
    \'%s\', 
    \'%s\')
'
  sql <- sprintf(sql_blanks, 
        "derivation", # actiontypecv
        "calc_sdg", # methodcode
        Sys.time(), # begindatetime
        utcoffset, # utcoffset
        sdg_results$hs[x], # samplingfeaturecode of sample
        UUIDgenerate(), #uuid
        "measurement", # resulttypecv
        4, # processinglevelid level 3 - analysis using other data
        "liquidAqueous",# sampledmediumcv
        1,
        UUIDgenerate(), #uuid
        "measurement", # resulttypecv
        4, # processinglevelid level 3 - analysis using other data
        "liquidAqueous",# sampledmediumcv
        1, # valuecount
        # methane results
        "notCensored", 
        sdg_results$qualitycode[x],
        "sporadic", 1, 1157, 
        sdg_results$dissolvedCH4_umol[x], sdg_results$valuedatetime[x], utcoffset,
        # carbon dioxide results
        "notCensored", 
        sdg_results$qualitycode[x],
        "sporadic", 1, 1157,
        sdg_results$dissolvedCO2_umol[x], sdg_results$valuedatetime[x], utcoffset)
  
  sql <- gsub("\n", "", sql)
  dbGetQuery(db, sql)
}

```

insert data
```{r, eval=FALSE}
db_insert_sdg_results(1) # one at a time

lapply(2:nrow(sdg_results), function(x) db_insert_sdg_results(x)) # or multiple

```

* add 2 results associated with sample analysis feature action - co2 umol per L and ch4 umol per L
* variable = methaneDissolved, carbonDioxideDissolved
* units = umol per L
* add measurement results 
* add measurement result values

**Query results from database**

```{r}
sql <- "SELECT mrv.datavalue, mrv.valuedatetime, sf.samplingfeaturecode, r.featureactionid, v.variablecode, u.unitsname, mr.qualitycodecv
  FROM odm2.measurementresultvalues mrv, odm2.results r, odm2.variables v, odm2.units u, odm2.samplingfeatures sf, odm2.featureactions fa, odm2.methods m, odm2.measurementresults mr
 WHERE r.variableid = v.variableid 
 AND r.featureactionid = fa.featureactionid
 AND fa.samplingfeatureid = sf.samplingfeatureid
 AND r.unitsid = u.unitsid
 AND mrv.resultid = r.resultid 
 AND mr.resultid = r.resultid 
 AND r.sampledmediumcv = 'liquidAqueous'
 AND m.methodcode = 'calc_sdg'"

sql <- gsub("\n", "", sql)
sdg_results_db <- dbGetQuery(db, sql)
```

then merge with samps join sites table

```{r}
sdg_results_db %>% 
  left_join(samps_join_sites, by = c("samplingfeaturecode" = "specimencode")) %>%
  rename(site = samplingfeaturecode.y) %>% kable()
```



add relationship between results to show that umol per L data are derived from ppm data of 2 samples, water temp, pressure data? 

# Miscellaneous

* Compare GSODR data for barometric pressure vs. field measurements
* Compare GSODR data for air temperature to field measurements of air and water temperature
