---
title: "Headspace sample processing Part 1"
output: 
  html_document:
    toc: yes
    toc_float:
      collapsed: false
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE, message=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(knitr)
library(readr)
library(dplyr)
library(RPostgreSQL)
library(uuid)
```

```{r connect, include=FALSE, message=FALSE}
password <- scan(".pgpass", what="")
db <- dbConnect(PostgreSQL(), 
                host = "sesync-postgis01.research.sesync.org",
                dbname = "choptank", 
                user = "palmergroup",
                password = password)
```

## Description

insert GC data into database

## Data

gc_results_db_test file should have all necessary sample information
add in site relationships with separate table of relationships (hasPart? isPartOf? isRelatedTo?)

```{r data, message=FALSE}
gc_data <- read_csv("data/gc_results_db_test.csv")
hs_sources <- read_csv("data/headspace_sources.csv")
```

```{r}
gc_for_db <- gc_data %>% 
  filter(Method %in% c("HS", "air")) %>%
  dplyr::select(Field_sample_date, Exetainer_ID, Site, Method, Time, ch4_ppm, co2_ppm, gc_date, gc_time)
```

## Database checks

Make sure all the metadata fields exist in the database

**Sites**

Compare the site names in the input data to the sampling feature codes in the database

```{r checksites}
dbsites <- dbGetQuery(db, "SELECT samplingfeaturecode FROM odm2.samplingfeatures WHERE samplingfeaturetypecv = 'site'")
sites <- unique(gc_for_db$Site)

data.frame(sites = sites,
           in_db = sites %in% dbsites$samplingfeaturecode) %>% 
  arrange(in_db) %>% kable()
```

add new sites to database if needed

```{r addsites, eval=FALSE}
db_add_site(samplingfeaturecode = "BBout", samplingfeaturedescription = "outflow of Bubbly Bay", samplingfeaturename = "BB out", samplingfeaturetypecv = "site")

db_add_site(samplingfeaturecode = "TBwell", samplingfeaturedescription = "TB center well", samplingfeaturename = "TB well", samplingfeaturetypecv = "site")

db_add_site(samplingfeaturecode = "TBin", samplingfeaturedescription = "presumed inflow of TB between ditch T and TB", samplingfeaturename = "TBin", samplingfeaturetypecv = "site")

db_add_site(samplingfeaturecode = "TBout", samplingfeaturedescription = "outflow of TB away from ditch T", samplingfeaturename = "TB out", samplingfeaturetypecv = "site")

```

**Method**

check for method
```{r checkmethods, eval=FALSE }
dbGetQuery(db, "SELECT * FROM odm2.methods")
```
```{r checkmethods2, eval=TRUE,echo=FALSE }
dbGetQuery(db, "SELECT * FROM odm2.methods") %>% kable()
```

Add new methods if they don't exist

```{r addmethods, eval=FALSE}
db_add_method(methodtypecv = "specimenCollection",
              methodcode = "HS",
              methodname = "Headspace dissolved gas",
              methoddescription = "Collect headspace dissolved gas sample")

db_add_method(methodtypecv = "specimenCollection",
              methodcode = "HSair",
              methodname = "Headspace source air",
              methoddescription = "Collect source air for headspace dissolved gas measurement")
```

**Variables**

check for variables

```{r checkvars, eval=FALSE}
dbGetQuery(db, "SELECT * FROM odm2.variables")
```
```{r checkvars2, echo=FALSE}
dbGetQuery(db, "SELECT * FROM odm2.variables") %>% kable()
```

add variables 
```{r addvars, eval=FALSE}
db_add_new_variable(variabletypecv = "WaterQuality", variablecode = "methane", variablenamecv = "Methane", variabledefinition = "Methane (CH4)")
db_add_new_variable(variabletypecv = "WaterQuality", variablecode = "carbonDioxide", variablenamecv = "Carbon dioxide", variabledefinition = "Carbon dioxide (CO2)")
```

**Units**

check for units

```{r checkunits, eval=FALSE}
dbGetQuery(db, "SELECT * FROM odm2.units WHERE unitsabbreviation = 'ppm'")
```
```{r checkunits2, include=FALSE}
dbGetQuery(db, "SELECT * FROM odm2.units WHERE unitsabbreviation = 'ppm'") %>% kable()
```

add necessary units if they aren't there

## Insert sample values

Each row in the `gc_for_db` table represents one **specimen** sampling feature that **wasCollectedAt** a wetland or stream sampling feature **site**

```{r}
head(gc_for_db) %>% kable()
gc_for_db$methodcode <- replace(gc_for_db$Method, which(gc_for_db$Method == "air"), "HSair")
```

**datetime formatting**

combine the Field sample date and Time columns into one

```{r}
gc_for_db$datetime <- strptime(paste(gc_for_db$Field_sample_date, gc_for_db$Time),
         format = "%d-%b-%y %H:%M:%S", tz = "")
gc_for_db$gc_datetime <- strptime(paste(gc_for_db$gc_date, gc_for_db$gc_time),
         format = "%m/%d/%y %H:%M:%S", tz = "")
```

**utc offset**

```{r}
utcoffset <- format(Sys.time(), "%z")
utcoffset <- as.integer(substr(utcoffset, 1,3))
```

```{r}
resulttypecv = "measurement"
censorcodecv = "notCensored"
qualitycodecv = "unknown"
aggregationstatisticcv = "sporadic"
timeaggregationinterval = 1 # assume 1 minute
timeaggregationintervalunitsid = 1157 # minute
processlinglevel = 2
valuecount = 1
```

## Insert new samples

add new samples, action for collecting that sample, relationship to site it was collected at

* add HS or air sample to sampling feature list as a specimen 
* for HS samples specimen medium is liquidAqueous
* for air samples specimen medium is gas
* related to site using wasCollectedAt relationship type
* add action for sample collection at date and time of field collection
* add feature action with new sample and new action for sample collection

```{r, include=FALSE, eval=FALSE}
sql_blanks <- 'WITH
  newsf AS (
  INSERT INTO odm2.samplingfeatures (
    samplingfeaturecode, samplingfeaturename, samplingfeaturetypecv, samplingfeatureuuid)
  VALUES (
    \'%s\', \'%s\', \'%s\', \'%s\')
  RETURNING samplingfeatureid),

  newrelation AS (
  INSERT INTO odm2.relatedfeatures (
    samplingfeatureid, relationshiptypecv, relatedfeatureid)
  VALUES (
    (SELECT newsf.samplinfeatureid FROM newsf),
    \'%s\',
    (SELECT samplingfeatureid FROM odm2.samplingfeatures WHERE samplingfeaturecode = \'%s\'))),

  newact AS (
  INSERT INTO odm2.actions (
    actiontypecv, methodid, begindatetime, begindatetimeutcoffset)
  VALUES (
    \'%s\', 
    (SELECT methodid FROM odm2.methods WHERE methodcode = \'%s\'), 
    \'%s\', 
    \'%s\')
  RETURNING actionid),

  newfa AS (
  INSERT into odm2.featureactions (
    samplingfeatureid, actionid)
  VALUES (
    (SELECT newsf.samplingfeatureid FROM newsf),
    (SELECT newact.actionid FROM newact))
  RETURNING featureactionid)'

```

generalize this with dataframe named samples and sample id column as sample id

```{r, include=FALSE, eval=FALSE}
x = 1
sql <- sprintf(sql_blanks, 
        gc_for_db$Exetainer_ID[x], # samplingfeaturecode
        paste(gc_for_db$methodcode[x], "sample", gc_for_db$Exetainer_ID[x]), # samplingfeaturename
        "specimen", # samplingfeaturetypecv
        UUIDgenerate(), # samplingfeatureuuid
        "wasCollectedAt", # relationshiptypecv
        gc_for_db$Site[x], # samplingfeaturecode (related feature)
        "specimenCollection", # actiontypecv
        gc_for_db$methodcode[x], # methodcode
        gc_for_db$datetime[x], # begindatetime
        utcoffset # utcoffset
        )
```

```{r, include=FALSE, eval=FALSE}
write(sql, "example_insert_query.txt")
```

```{r, include=FALSE, eval=FALSE}
sql <- gsub("\n", "", sql)
```

**Function to insert new samples**

```{r define_db_insert_samples}
db_insert_samples <- function(x){
  sql_blanks <- 'WITH
  newsf AS (
  INSERT INTO odm2.samplingfeatures (
    samplingfeaturecode, samplingfeaturename, samplingfeaturetypecv, samplingfeatureuuid)
  VALUES (
    \'%s\', \'%s\', \'%s\', \'%s\')
  RETURNING samplingfeatureid),

  newrelation AS (
  INSERT INTO odm2.relatedfeatures (
    samplingfeatureid, relationshiptypecv, relatedfeatureid)
  VALUES (
    (SELECT newsf.samplingfeatureid FROM newsf),
    \'%s\',
    (SELECT samplingfeatureid FROM odm2.samplingfeatures WHERE samplingfeaturecode = \'%s\'))),

  newact AS (
  INSERT INTO odm2.actions (
    actiontypecv, methodid, begindatetime, begindatetimeutcoffset)
  VALUES (
    \'%s\', 
    (SELECT methodid FROM odm2.methods WHERE methodcode = \'%s\'), 
    \'%s\', 
    \'%s\')
  RETURNING actionid)

  INSERT INTO odm2.featureactions (
    samplingfeatureid, actionid)
  VALUES (
    (SELECT newsf.samplingfeatureid FROM newsf),
    (SELECT newact.actionid FROM newact))
  RETURNING featureactionid'
  
  sql <- sprintf(sql_blanks, 
        gc_for_db$Exetainer_ID[x], # samplingfeaturecode
        paste(gc_for_db$methodcode[x], "sample", gc_for_db$Exetainer_ID[x]), # samplingfeaturename
        "specimen", # samplingfeaturetypecv
        UUIDgenerate(), # samplingfeatureuuid
        "wasCollectedAt", # relationshiptypecv
        gc_for_db$Site[x], # samplingfeaturecode (related feature)
        "specimenCollection", # actiontypecv
        gc_for_db$methodcode[x], # methodcode
        gc_for_db$datetime[x], # begindatetime
        utcoffset # utcoffset
        )
  sql <- gsub("\n", "", sql)
  dbGetQuery(db, sql)
}
```

Run function on new dataset - edit this so it just accepts the dataframe

```{r insertsamples, eval=FALSE}
db_insert_samples(1) # one at a time
lapply(2:nrow(gc_for_db), function(x) db_insert_samples(x)) # or multiple
```

Get feature action with information

```{r}
db_get_featureaction <- function(featureactionid = 309){
  sql <- sprintf("SELECT fa.featureactionid, sf.samplingfeaturecode, act.actiontypecv, act.begindatetime, m.methodcode 
                 FROM odm2.featureactions fa, odm2.samplingfeatures sf, odm2.actions act, odm2.methods m
                 WHERE fa.samplingfeatureid = sf.samplingfeatureid
                 AND fa.actionid = act.actionid
                 AND act.methodid = m.methodid
                 AND fa.featureactionid = '%s'", featureactionid)
  
  sql <- gsub("\n", "", sql)
  dbGetQuery(db, sql)
}
```

```{r}
db_get_featureaction(341) %>% kable()
```


## Add source air relationships

* add relationships between source gas and hs samples

Not sure what this type of relationship should be. Source gas for headspace samples is part of the headspace sample, and then it is corrected for later in the calculation of dissolved gas in the original water sample. (Should probably update method description to make a note of the sample being diluted by the concentration of the source air, with specific volumes of the sample and source air.)

For now, let's say the source gas sample **isPartOf** the headspace gas sample. 

```{r}
head(hs_sources) %>% kable()
```

Join the sampling features table from the database to the dataframe with the relationships to get the IDs. This is a sequence of two joins to get the ids for the child and parent sampling features. These joins could also be done on the SQL side.

Get sampling ids from the samplingfeatures table

```{r}
specimens <- dbGetQuery(db, "SELECT samplingfeatureid, samplingfeaturecode FROM odm2.samplingfeatures WHERE samplingfeaturetypecv = 'specimen'")
```


```{r}
new_related_samples <- hs_sources %>%  
  left_join(specimens, by = c("headspace_sample_id" = "samplingfeaturecode")) %>% 
  rename(relatedfeatureid = samplingfeatureid) %>%
  mutate(relationshiptypecv = "isPartOf") %>%
  left_join(specimens, by = c("source_gas_sample_id" = "samplingfeaturecode")) %>%
  dplyr::select(samplingfeatureid, relationshiptypecv, relatedfeatureid)
```

Update table in database

```{r, eval= FALSE}
dbWriteTable(db, c("odm2","relatedfeatures"),
             new_related_samples,
             overwrite = FALSE,
             row.names = FALSE,
             append = TRUE)
```

Get all headspace samples and paired source air samples

```{r}
db_get_hs_sources <- function(){
  sql <- "WITH t1 AS(
    SELECT 
      REL.samplingfeatureid,
      SF.samplingfeaturecode,
      REL.relationshiptypecv,
      REL.relatedfeatureid
    FROM 
     odm2.relatedfeatures REL,
     odm2.samplingfeatures SF
    WHERE 
     REL.samplingfeatureid = SF.samplingfeatureid
     AND REL.relationshiptypecv = 'isPartOf')

    SELECT
      t1.samplingfeaturecode source,
      SF2.samplingfeaturecode HS
    FROM
      t1,
      odm2.samplingfeatures SF2
    WHERE
      t1.relatedfeatureid = SF2.samplingfeatureid
  "
sql <- gsub("\n", " ", sql)

dbGetQuery(db, sql)  
}

```

```{r}
db_get_hs_sources() %>% kable()
```

## Insert sample PPM data

* add action for specimen analysis 
* add feature action with new sample and new action for sample analysis
* add 2 results associated with sample analysis feature action - co2 ppm and ch4 ppm
* variable = methane (methaneDissolved comes later), carbonDioxide
* units = ppm, Part per Million
* add measurement results 
* add measurement result values

**Add method for GC analysis**

```{r}
db_add_method(methodtypecv = "specimenAnalysis",
              methodcode = "GC",
              methodname = "Gas chromatography",
              methoddescription = "Analyze gas componds on gas chromatograph")
```


```{r}
sql_blanks <- 'WITH
  newact AS (
  INSERT INTO odm2.actions (
    actiontypecv, methodid, begindatetime, begindatetimeutcoffset)
  VALUES (
    \'%s\', 
    (SELECT methodid FROM odm2.methods WHERE methodcode = \'%s\'), 
    \'%s\', 
    \'%s\')
  RETURNING actionid),

  newfa AS (
  INSERT into odm2.featureactions (
    samplingfeatureid, actionid)
  VALUES (
    (SELECT samplingfeatureid FROM odm2.samplingfeatures WHERE samplingfeaturecode = \'%s\'),
    (SELECT newact.actionid FROM newact))
  RETURNING featureactionid),

  newresult AS (
  INSERT INTO odm2.results (
    featureactionid, resultuuid, resulttypecv, variableid, unitsid, processinglevelid, sampledmediumcv, valuecount)
  VALUES (
    
    (SELECT newfa.featureactionid FROM newfa),
    \'%s\',
    \'%s\',
    (SELECT variableid FROM odm2.variables WHERE variablecode = \'methane\'),
    (SELECT unitsid FROM odm2.units WHERE unitsabbreviation = \'ppm\'),
    \'%s\',
    \'%s\',
    \'%s\'),
  (
    (SELECT newfa.featureactionid FROM newfa),
    \'%s\',
    \'%s\',
    (SELECT variableid FROM odm2.variables WHERE variablecode = \'carbonDioxide\'),
    (SELECT unitsid FROM odm2.units WHERE unitsabbreviation = \'ppm\'),
    \'%s\',
    \'%s\',
    \'%s\')
  RETURNING resultid, variableid),

  newmr1 AS (
  INSERT INTO odm2.measurementresults (
    resultid,
    censorcodecv,
    qualitycodecv,
    aggregationstatisticcv,
    timeaggregationinterval,
    timeaggregationintervalunitsid)
  VALUES (
    (SELECT newresult.resultid FROM newresult WHERE variableid = (SELECT variableid FROM odm2.variables WHERE variablecode = \'methane\')),
    \'%s\',
    \'%s\', 
    \'%s\', 
    \'%s\', 
    \'%s\')),

  newmrv1 AS (
  INSERT INTO odm2.measurementresultvalues (
    resultid,
    datavalue,
    valuedatetime,
    valuedatetimeutcoffset)
  VALUES (
    (SELECT newresult.resultid FROM newresult WHERE variableid = (SELECT variableid FROM odm2.variables WHERE variablecode = \'methane\')),
    \'%s\',
    \'%s\', 
    \'%s\')),

  newmr3 AS (
  INSERT INTO odm2.measurementresults (
    resultid,
    censorcodecv,
    qualitycodecv,
    aggregationstatisticcv,
    timeaggregationinterval,
    timeaggregationintervalunitsid)
  VALUES (
    (SELECT newresult.resultid FROM newresult WHERE variableid = (SELECT variableid FROM odm2.variables WHERE variablecode = \'carbonDioxide\')),
    \'%s\',
    \'%s\', 
    \'%s\', 
    \'%s\', 
    \'%s\'))

  INSERT INTO odm2.measurementresultvalues (
    resultid,
    datavalue,
    valuedatetime,
    valuedatetimeutcoffset)
  VALUES (
    (SELECT newresult.resultid FROM newresult WHERE variableid = (SELECT variableid FROM odm2.variables WHERE variablecode = \'carbonDioxide\')),
    \'%s\',
    \'%s\', 
    \'%s\')
    
'

```

Not sure if sampled medium should be liquidAqueous for HS samples and gas for air samples, or gas for both. Using gas for both now. 

Date time of measurement result values for ppm measurements on GC is linked back to field sampling datetime for now, even though that is the specimen collection action not the specimen analysis action, because the date time of the GC run is not important except for merging data.

```{r}
db_insert_gc_results <- function(x){
  sql_blanks <- 'WITH
  newact AS (
  INSERT INTO odm2.actions (
    actiontypecv, methodid, begindatetime, begindatetimeutcoffset)
  VALUES (
    \'%s\', 
    (SELECT methodid FROM odm2.methods WHERE methodcode = \'%s\'), 
    \'%s\', 
    \'%s\')
  RETURNING actionid),

  newfa AS (
  INSERT into odm2.featureactions (
    samplingfeatureid, actionid)
  VALUES (
    (SELECT samplingfeatureid FROM odm2.samplingfeatures WHERE samplingfeaturecode = \'%s\'),
    (SELECT newact.actionid FROM newact))
  RETURNING featureactionid),

  newresult AS (
  INSERT INTO odm2.results (
    featureactionid, resultuuid, resulttypecv, variableid, unitsid, processinglevelid, sampledmediumcv, valuecount)
  VALUES (
    
    (SELECT newfa.featureactionid FROM newfa),
    \'%s\',
    \'%s\',
    (SELECT variableid FROM odm2.variables WHERE variablecode = \'methane\'),
    (SELECT unitsid FROM odm2.units WHERE unitsabbreviation = \'ppm\'),
    \'%s\',
    \'%s\',
    \'%s\'),
  (
    (SELECT newfa.featureactionid FROM newfa),
    \'%s\',
    \'%s\',
    (SELECT variableid FROM odm2.variables WHERE variablecode = \'carbonDioxide\'),
    (SELECT unitsid FROM odm2.units WHERE unitsabbreviation = \'ppm\'),
    \'%s\',
    \'%s\',
    \'%s\')
  RETURNING resultid, variableid),

  newmr1 AS (
  INSERT INTO odm2.measurementresults (
    resultid,
    censorcodecv,
    qualitycodecv,
    aggregationstatisticcv,
    timeaggregationinterval,
    timeaggregationintervalunitsid)
  VALUES (
    (SELECT newresult.resultid FROM newresult WHERE variableid = (SELECT variableid FROM odm2.variables WHERE variablecode = \'methane\')),
    \'%s\',
    \'%s\', 
    \'%s\', 
    \'%s\', 
    \'%s\')),

  newmrv1 AS (
  INSERT INTO odm2.measurementresultvalues (
    resultid,
    datavalue,
    valuedatetime,
    valuedatetimeutcoffset)
  VALUES (
    (SELECT newresult.resultid FROM newresult WHERE variableid = (SELECT variableid FROM odm2.variables WHERE variablecode = \'methane\')),
    \'%s\',
    \'%s\', 
    \'%s\')),

  newmr3 AS (
  INSERT INTO odm2.measurementresults (
    resultid,
    censorcodecv,
    qualitycodecv,
    aggregationstatisticcv,
    timeaggregationinterval,
    timeaggregationintervalunitsid)
  VALUES (
    (SELECT newresult.resultid FROM newresult WHERE variableid = (SELECT variableid FROM odm2.variables WHERE variablecode = \'carbonDioxide\')),
    \'%s\',
    \'%s\', 
    \'%s\', 
    \'%s\', 
    \'%s\'))

  INSERT INTO odm2.measurementresultvalues (
    resultid,
    datavalue,
    valuedatetime,
    valuedatetimeutcoffset)
  VALUES (
    (SELECT newresult.resultid FROM newresult WHERE variableid = (SELECT variableid FROM odm2.variables WHERE variablecode = \'carbonDioxide\')),
    \'%s\',
    \'%s\', 
    \'%s\')
    
'
  sql <- sprintf(sql_blanks, 
        "specimenAnalysis", # actiontypecv
        "GC", # methodcode
        gc_for_db$gc_datetime[x], # begindatetime
        utcoffset, # utcoffset
        gc_for_db$Exetainer_ID[x], # samplingfeaturecode of sample
        UUIDgenerate(), #uuid
        "measurement", # resulttypecv
        3, # processinglevelid level 2 - derived from calibration curve
        "gas",# sampledmediumcv
        1,
        UUIDgenerate(), #uuid
        "measurement", # resulttypecv
        3, # processinglevelid level 2 - derived from calibration curve
        "gas",# sampledmediumcv
        1, # valuecount# valuecount
        # methane results
        "notCensored", "unknown", "sporadic", 1, 1157, 
        gc_for_db$ch4_ppm[x], gc_for_db$datetime[x], utcoffset,
        # carbon dioxide results
        "notCensored", "unknown", "sporadic", 1, 1157,
        gc_for_db$co2_ppm[x], gc_for_db$datetime[x], utcoffset
        )
  
  sql <- gsub("\n", "", sql)
  dbGetQuery(db, sql)
}
  
```

insert results. Doesn't seem to work with NAs

```{r}
db_insert_gc_results(1) # one at a time

lapply(2:nrow(gc_for_db), function(x) db_insert_gc_results(x)) # or multiple

```

# Query

try to get ppm data for TBair KLH 521

```{r}
sql <- "SELECT mrv.datavalue, mrv.valuedatetime, sf.samplingfeaturecode, r.featureactionid, v.variablecode, u.unitsname 
  FROM odm2.measurementresultvalues mrv, odm2.results r, odm2.variables v, odm2.units u, odm2.samplingfeatures sf, odm2.featureactions fa
 WHERE r.variableid = v.variableid 
 AND r.featureactionid = fa.featureactionid
 AND fa.samplingfeatureid = sf.samplingfeatureid
 AND r.unitsid = u.unitsid
 AND mrv.resultid = r.resultid 
 AND r.sampledmediumcv = 'gas'
 AND sf.samplingfeaturecode = 'KLH 521'"

sql <- gsub("\n", "", sql)
dbGetQuery(db, sql)
```

