---
title: "Headspace sample processing Part 2"
output: 
  html_document:
    toc: yes
    toc_float:
      collapsed: false
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE, message=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(knitr)
library(readr)
library(dplyr)
library(RPostgreSQL)
library(uuid)
```

```{r connect, include=FALSE, message=FALSE}
password <- scan(".pgpass", what="")
db <- dbConnect(PostgreSQL(), 
                host = "sesync-postgis01.research.sesync.org",
                dbname = "choptank", 
                user = "palmergroup",
                password = password)
```

## Description

insert dissolved gas concentration data into database

## Data

* Get PPM data from database for source air and samples
* ppmv data for source gas - determined from GC workflow
* ppmv data for sample gas - determined from GC workflow
* water temperature - measured in the field or offset from GSODR
* barometric pressure - measured in the field or from GSODR data
* source gas sample volume - default 20 mL
* sample gas sample volume - default 40 mL

Reshape the data so there is one row for each headspace sample with the required inputs for the `def.calc.sdg()` function. 

# Calculations

Calculations are based on the NEON dissolved gas function. Add proper citation [here](). 

```{r}
source('def.calc.sdg.R')
```

Use function by supplying column names and defaults. Output is concentrations and 100% saturation concentrations as additional columns on the data frame. 

```{r, eval=FALSE}
sdg_out <- def.calc.sdg(inputFile = sdg_data, 
                        volGas = 20,
                        volH2O = 40,
                        baro = STP_kPa,
                        waterTemp = waterTemp,
                        headspaceTemp = waterTemp,
                        eqCO2 = ppm_CO2_sample,
                        sourceCO2 = ppm_CO2_source,
                        eqCH4 = ppm_CH4_sample,
                        sourceCH4 = ppm_CH4_source)
```

Reshape concentration data to put values in database for each sample

# Insert concentration results

* add action for specimen analysis(?)
* add feature action with new sample and new action for sample analysis
* add 2 results associated with sample analysis feature action - co2 umol per L and ch4 umol per L
* variable = methaneDissolved, carbonDioxideDissolved
* units = umol per L
* add measurement results 
* add measurement result values

add relationship between results to show that umol per L data are derived from ppm data of 2 samples, water temp, pressure data? 

# Miscellaneous

* Compare GSODR data for barometric pressure vs. field measurements
* Compare GSODR data for air temperature to field measurements of air and water temperature
